---
output: github_document
---

# plinko: Animated Plinko Boards

<!-- badges: start -->
[![R build status](https://github.com/mjskay/plinko/workflows/R-CMD-check/badge.svg)](https://github.com/mjskay/plinko/actions)
[![Coverage status](https://codecov.io/gh/mjskay/plinko/branch/master/graph/badge.svg)](https://codecov.io/github/mjskay/plinko?branch=master)
<!-- badges: end -->

`plinko` is an R package for creating animated Plinko boards (as known as Galton 
boards, quincunxes, or bean machines). It helps determine Plinko board parameters
and generates plausible paths through a Plinko board that result in a desired
sample or set of quantiles, then renders the Plinko board as an animated GIF or 
video.

## Installation

You can install plinko with:

```r
# install.packages("remotes")
remotes::install_github("mjskay/plinko")
```

## Example

This example requires the following libraries:

```{r setup, message=FALSE, warning=FALSE}
library(plinko)
library(ggplot2)
library(distributional)
library(ggdist)

theme_set(theme_ggdist())
```

You can construct plinko boards using distributions from the 
[distributional](https://pkg.mitchelloharawild.com/distributional/) package:

```{r}
set.seed(1234)  # for reproducibility

dist = dist_normal(1,2)
board = plinko_board(dist, n_bin = 10, n_ball = 50)
autoplot(board)
```


