---
output: github_document
---

# plinko: Animated Plinko Boards

<!-- badges: start -->
[![R build status](https://github.com/mjskay/plinko/workflows/R-CMD-check/badge.svg)](https://github.com/mjskay/plinko/actions)
[![Coverage status](https://codecov.io/gh/mjskay/plinko/branch/master/graph/badge.svg)](https://codecov.io/github/mjskay/plinko?branch=master)
<!-- badges: end -->

`plinko` is an R package for creating animated Plinko boards (as known as Galton 
boards, quincunxes, or bean machines). It helps determine Plinko board parameters
and generates plausible paths through a Plinko board that result in a desired
sample or set of quantiles, then renders the Plinko board as an animated GIF or 
video.

## Installation

You can install plinko with:

```r
# install.packages("remotes")
remotes::install_github("mjskay/plinko")
```

## Example

This example requires the following libraries:

```{r setup, message=FALSE, warning=FALSE}
library(plinko)
library(dplyr)
library(ggplot2)
library(distributional)
library(ggdist)

theme_set(theme_ggdist())
```

You can construct plinko boards using distributions from the 
[distributional](https://pkg.mitchelloharawild.com/distributional/) package:

```{r board_plot, fig.width = 1, fig.height = 4}
set.seed(1234)  # for reproducibility

dist = dist_normal(2, 1)
board = plinko_board(dist, n_bin = 5, n_ball = 20)
autoplot(board)
```

The plot shows the pins and slots in the board, as well as the target distribution
(in orange), the binomial distribution used to approximate it (in black), and
20 quantiles from the target distribution (in blue). By default, quantiles of the
target distribution (rather than random samples) are used in order to ensure
the final distribution of balls is representative of the target distribution.
If you want a random sample of balls instead, you can pass `sampling = "random"`
to `plinko_board()`.

Once you are happy with the appearance of your plinko board, you can animate it
using `animate()`, which by default will omit the annotations of the paths 
and distributions (though these can be added back in with `show_paths = TRUE`, 
`show_dist = TRUE`, or `show_target_dist = TRUE`).
Here we'll also use `tween_balls()` to add some tweening frames so the balls drop
more smoothly, and set a `height` on the animation (given a desired height,
`animate.plinko_board()` will attempt to find a reasonable `width` automatically):

```{r board_animation}
board %>%
  tween_balls(frame_mult = 4) %>%
  animate(height = 450, fps = 30, end_pause = 2 * 30)
```

For a more comprehnsive example of how to customize your animation (including
customizing aesthetics, geoms, and annotations), see `vignette("presidential-plinko")`.
