---
title: "Presidential Plinko"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{presidential_plinko}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The following code constructs a Plinko board for a quantile dotplot of a
predictive distribution for the 2020 US Presidential election outcome.

See [presidential-plinko.com](https://presidential-plinko.com) and the 
[mjskay/election-galton-board](https://github.com/mjskay/election-galton-board)
repository for more information.

## Setup

The following packages are required for this example:

```{r setup}
library(plinko)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(ggdist)
library(ggforce)
library(tweenr)
library(ragg)
library(snakecase)
library(gifski)

theme_set(theme_ggdist())
```

## Desired target sample

Say we've taken 50 quantiles from a predictive distribution for the number of 
electoral votes Biden will receive in the 2020 US Presidential election:

```{r}
# this is based on the Economist's forecast as of Sept 27, 2020.
x = c(200, 217, 228, 239, 249, 256, 264, 270, 274, 279, 285, 290,  292, 296, 302, 306, 308, 310, 316, 319, 319, 319, 323, 325, 329,  334, 335, 338, 340, 343, 347, 350, 352, 357, 357, 358, 361, 363,  368, 372, 374, 374, 377, 383, 388, 394, 399, 412, 415, 427)

tibble(x) %>%
  ggplot(aes(x = x)) +
  geom_dots() +
  xlab("Predicted electoral votes for Biden")
```

What we would like to do is construct a Plinko board that "sediments" into this dotplot. That is,
it gives us an animation of a plausible random process that could have led to a dotplot
that looks like the one above. This will take several steps.

## Constructing the board

First, to determine the board dimensions, we need to decide on the number of draws we will 
take and the number of pins in one row of the board (equivalently, the size of the Binomial 
distribution we are simulating with the board):

```{r}
n_bin = 8
bin_width = 38
```

Then we can construct the Plinko board:

```{r}
set.seed(12345)
board = plinko_board(
  x, n_bin, bin_width, center = 326,
  n_ball = 1,
  limits = c(0, 538)#, total_height = 2445
)
board
```

## Plotting the board

The Plinko board will automatically find random paths the balls could have taken
to end up in their final locations. We can see these paths by plotting the board
board with the `plot()` function, which returns a `ggplot()` object:

```{r}
plot(board)
```

By default the plot overlays the Binomial distribution implied by the Plinko board so 
you can judge the quality of the approximation. We can also turn this off:

```{r}
plot(board, show_dist = FALSE)
```

We could also create a plot like the one above manually by using the `slot_edges()`, `pins()`, `paths()`, and `balls()`
unctions, which return data frames containing the locations of all of the elements of the board:

```{r board_plot}
ggplot() + 
  geom_segment(aes(x = x, y = 0, xend = x, yend = height), data = slot_edges(board), color = "gray75", size = 1) + 
  geom_point(aes(x, y), data = pins(board), shape = 19, color = "#e41a1c", size = 1) +
  geom_path(aes(x = x, y = y, group = ball_id), data = paths(board), alpha = 1/4, size = 1, color = "gray50") +
  geom_circle(aes(x0 = x, y0 = y, r = width/2), data = balls(board), fill = "#1f78b4", color = NA) +
  coord_fixed(expand = FALSE, clip = "off") +
  ylab("") +
  scale_y_continuous(breaks = NULL) +
  theme(
    axis.line.y = element_blank(),
    axis.line.x = element_line(color = "gray75", size = 1)
  ) 
```

You can also see the board without paths by passing `show_paths = FALSE`, and you 
can plot specific frames of the animation by passing the `frame` parameter:

```{r}
plot(board, show_paths = FALSE, frame = 9)
```



## Animating the board

By default, no tweening is done between balls on the board:

```{r}
animate(board, fps = 7.5)
```

Most likely you will want some tweening between ball states. I find that using 4 times the
base number of frames with a `"bounce-out"` easing function makes for a smooth
animation that feels more physically accurate. You can add tweening using the
`tween_balls()` function:

```{r}
board %>%
  tween_balls(frame_mult = 4, ease = "bounce-out") %>%
  animate(fps = 30)
```


First, we need to set up the output folder to hold the PNG files that will contain the frames:

```{r png_temp}
png_dir = "png_temp"
dir.create(png_dir, showWarnings = FALSE)
unlink(list.files(png_dir, pattern = "*.png", full.names = TRUE), expand = FALSE)
```

Then, we render all frame of the Galton board one at a time into PNG files in the above directory.

```{r render_frames, message=FALSE, warning=FALSE, results=FALSE}
x0 = seq(-n_bin/2, n_bin/2) * bin_width + center
x0 = x0[min(slot_edges) < x0 & x0 < max(slot_edges)]

# figure out aspect ratio
height_px = 800
width_px = (x_max - x_min + 90)/total_height * height_px

Biden_color = "#0571b0"
Trump_color = "#ca0020"

tweened_frames = group_split(tweened_frames_df, frame_id)
# repeat the last frame to match up to the desired exact length of the animation
last_frame = tweened_frames[length(tweened_frames)]
tweened_frames = c(tweened_frames, rep(last_frame, n_frame * frame_mult - length(tweened_frames)))

pb = txtProgressBar(max = length(tweened_frames), style = 3, file = stderr())
for (i in seq_along(tweened_frames)) {
  tweened_frame = tweened_frames[[i]]
  outfile <- sprintf("%s/%04i.png", png_dir, i)
  agg_png(outfile, width = width_px, height = height_px, res = 100)
  
  p = ggplot(tweened_frame) + 
    geom_point(aes(x, y), size = 1, shape = 18, color = "gray50", data = pins_df) +
    geom_segment(
      aes(x, 0, xend = x, yend = height),
      size = 1, color = "gray75", data = slots_df
    ) +
    geom_circle(
      aes(
        x0 = x,
        y0 = y,
        r = ball_width/2,
        fill = ifelse(y > slot_height, "none", ifelse(x <= 269, "Trump", "Biden")),
        group = ball_id
      ), color = NA) + 
    (if (show_dist) geom_step(
      data = data.frame(x0),
      aes(
        x = x0,
        y = n_ball * ball_width * dbinom(
          round((x0 - center)/bin_width + n_bin/2),
          size = n_bin,
          prob = 0.5
        )),
      col="black",
      alpha = 0.2,
      size = 1,
      direction = "mid"
    )) +
    geom_vline(xintercept = 269, color = "black", alpha = 0.15, size = 1) +
    annotate("text", 
      x = 290, y = 0.95 * total_height, 
      label = "Biden wins", hjust = 0, color = Biden_color
      # fontface = "bold",
    ) +
    annotate("text", 
      x = 250, y = 0.95 * total_height,
      label = "Trump wins", hjust = 1, color = Trump_color
      # fontface = "bold",
    ) +
    annotate("label", 
      x = 269, y = 0.98 * total_height,
      label = "269", hjust = 0.5, color = "gray50",
      fontface = "bold"
    ) +
    scale_fill_manual(
      limits = c("none", "Biden", "Trump"),
      values = c("gray45", Biden_color, Trump_color), 
      guide = FALSE
    ) +
    coord_fixed(ylim = c(0, total_height), xlim = c(x_min - 45, x_max + 45), 
      expand = FALSE, clip = "off") + 
    theme_ggdist() + 
    theme(
      axis.line.y = element_blank(),
      axis.line.x = element_line(color = "gray75", size = 1),
      axis.title.x = element_text(hjust = 0, size = 10, color = "gray25"),
      plot.title = element_text(hjust = 0.5),
      plot.title.position = "plot"
    ) +
    scale_y_continuous(breaks = NULL) +
    scale_x_continuous(limits = range(slot_edges, x_min, x_max)) +
    labs(
      x = paste0("Electoral votes for Biden\n\nPresidential Plinko, ", format(Sys.Date(), "%b %d"), "\n", 
        model_name, "'s model\n"),
      y = NULL
    ) +
    ggtitle(model_name)
  print(p)
  invisible(dev.off())

  setTxtProgressBar(pb, i)
}
```

Finally, we use `gifski()` to roll all the frames into an animated GIF:

```{r render_gif, message=FALSE, warning=FALSE, results=FALSE}
gif_name = paste0("galton_board-", to_snake_case(model_name), ".gif")

gifski(
  list.files(png_dir, pattern = "*.png", full.names = TRUE), 
  gif_name, 
  width = width_px,
  height = height_px,
  delay = 1/30,
  loop = TRUE,
  progress = TRUE
)
```

```{r echo=FALSE, results="asis"}
cat("![](", gif_name, ")", sep = "")
```
