---
title: "Presidential Plinko"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{presidential_plinko}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style type="text/css">
.kable-table table {
  margin-left: 0;
}
img {
  border: none;
}
</style>

```{r chunk_options, include=FALSE}
if (capabilities("cairo") && Sys.info()[['sysname']] != "Darwin") {
  knitr::opts_chunk$set(
    dev.args = list(png = list(type = "cairo"))
  )
}
```


The following code constructs a Plinko board for a quantile dotplot of a
predictive distribution for the 2020 US Presidential election outcome.

See [presidential-plinko.com](https://presidential-plinko.com) and the 
[mjskay/election-galton-board](https://github.com/mjskay/election-galton-board)
repository for more information.

## Setup

The following packages are required for this example:

```{r setup, message=FALSE, warning=FALSE}
library(plinko)
library(dplyr)
library(ggplot2)
library(ggdist)
library(ggforce)
library(distributional)

theme_set(theme_ggdist())
```

## Desired target sample

Say we want to display FiveThirtyEight's probabilistic prediction of Biden's chance
of winning the 2020 US Presidential election. The following dataset is FiveThirtyEight's
prediction as of Oct 16, 2020:

```{r pres_pred_2020_plot, fig.width = 6, fig.height = 4.5}
data(pres_pred_2020, package = "plinko")

pres_pred_2020 %>%
  ggplot(aes(x = total_ev, y = evprob_chal)) +
  geom_col(fill = "gray75") +
  geom_vline(xintercept = 269) +
  coord_cartesian(xlim = c(0, 538)) +
  labs(
    x = "Electoral votes for Biden",
    y = "Predicted probability",
    title = "Predictive distribution for Biden's electoral votes",
    subtitle = "FiveThirtyEight's model of the 2020 US election as of Oct 16, 2020"
  )  
```

This particular dataset is already summarized as a histogram: the `evprob_chal`
column is the probability associated with each x value (`total_ev`) in the
above chart. While we could operate directly on this summarized data using 
weighted statistics (weighted means, quantiles, etc)---and this is what I did
for [the implementation of presidential-plinko.com](https://github.com/mjskay/election-galton-board)---for
the purposes of this example, we will convert the summarized data into a sample
first. This type of data is more commonly encountered and a bit easier to work with.

To convert the data into a sample of (say) 5000, we can just use `sample()` and
provide the probabilities as weights:

```{r ev_sample}
set.seed(1234) # for reproducibility
ev_sample = sample(pres_pred_2020$total_ev, 5000, replace = TRUE, prob = pres_pred_2020$evprob_chal)
```

We can verify the data looks essential the same (down to sampling error):

```{r ev_sample_plot, fig.width = 6, fig.height = 4.5}
tibble(ev_sample) %>%
  ggplot(aes(x = ev_sample)) +
  geom_histogram(binwidth = 1, fill = "gray75") +
  geom_vline(xintercept = 269) +
  coord_cartesian(xlim = c(0, 538)) +
  labs(
    x = "Electoral votes for Biden",
    y = "Frequency",
    title = "Predictive distribution for Biden's electoral votes (sampled)",
    subtitle = "FiveThirtyEight's model of the 2020 US election as of Oct 16, 2020"
  )  
```

## Constructing a basic board

The `plinko_board()` function can be called directly on a numeric vector; however,
this usage typically requires you to already know exactly the board parameters
you need (number of bins, bin width, etc). Thus, it is usally easier to call
`plinko_board()` on a [distributional](https://pkg.mitchelloharawild.com/distributional/)
object, which is an object that represents some probability distribution. 

Given a distributional object and either a desired bin width (`bin_width`) **or**
number of bins (`n_bin`), `plinko_board()` will do its best to automatically figure 
out the necessary board parameters to accomodate some number of balls from that
distribution (`n_ball`, which by default is `50`).

We can construct a distributional object from sample data using `distributional::dist_sample()`
and pass it to `plinko_board()`:

```{r board_1}
ev_dist = dist_sample(list(ev_sample))
board = plinko_board(ev_dist, bin_width = 35)
board
```

## Plotting the board

### Using `autoplot()`

The Plinko board will automatically find random paths the balls could have taken
to end up in their final locations. We can see these paths by plotting the board
with the `autoplot()` function, which returns a `ggplot()` object:

```{r fig.height = 5, fig.width = 1.5}
board %>%
  autoplot()
```

By default the plot overlays the Binomial distribution implied by the Plinko board so 
you can judge the quality of the approximation. We can also turn this off:

```{r fig.height = 5, fig.width = 1.5}
board %>%
  autoplot(show_dist = FALSE)
```

You can also see the board without paths by passing `show_paths = FALSE`, and you 
can plot specific frames of the animation by passing the `frame` parameter:

```{r fig.height = 5, fig.width = 1.5}
autoplot(board, show_paths = FALSE, frame = 9)
```

### Plotting manually

We could also create a plot like the ones above manually by using the `slot_edges()`, `pins()`, `paths()`, and `balls()`
unctions, which return data frames containing the locations of all of the elements of the board:

```{r board_plot, fig.height = 5, fig.width = 1.5}
ggplot() + 
  geom_segment(aes(x = x, y = 0, xend = x, yend = height), data = slot_edges(board), color = "gray75", size = 1) + 
  geom_point(aes(x, y), data = pins(board), shape = 19, color = "#e41a1c", size = 1) +
  geom_path(aes(x = x, y = y, group = ball_id), data = paths(board), alpha = 1/4, size = 1, color = "gray50") +
  geom_circle(aes(x0 = x, y0 = y, r = width/2), data = balls(board), fill = "#1f78b4", color = NA) +
  coord_fixed(expand = FALSE, clip = "off") +
  ylab(NULL) +
  scale_y_continuous(breaks = NULL) +
  theme(
    axis.line.y = element_blank(),
    axis.line.x = element_line(color = "gray75", size = 1)
  ) 
```

However, if you wish to customize the plot, it is probably easier to use the
`modify_layer()` and `add_layers()` functions described later in this document, as
these also impact the frames used when an animated plot is rendered.

## Forcing a slot edge to fall on 269

Before we get to animating this board, we also want to address an additional 
constraint not handled automatically above: we will want to be able to color the
balls red if they fall below `269` (Trump wins) and blue if they fall above `269`
(Biden wins). To do that, we need a slot edge to fall on `269`.

Let's start by seeing
how well our current approximation does by turning off the display of ball paths
and adding on a vertical line at `269`. `autoplot()` returns a `ggplot` object
so this is straightforward:

```{r board_1_edge_plot, fig.height = 5, fig.width = 1.5}
autoplot(board, show_paths = FALSE) +
  geom_vline(xintercept = 269)
```

The slot edge closest to 269 is a little too far away for my taste. We can 
see how close it is in electoral votes:

```{r board_1_269_dist}
min(abs(board$slot_edges - 269))
```

A simple approach to fixing this problem might be to take a range of bin sizes we
are happy with (say 35 to 45 electoral votes) and find a bin width in that range 
that minimizes the above number:

```{r find_bin_width}
closest_edge_to_269 = Vectorize(function(bin_width) {
  board = plinko_board(ev_dist, bin_width = bin_width)
  min(abs(board$slot_edges - 269))
})

bin_width = (35:45)[which.min(closest_edge_to_269(35:45))]
bin_width
```

Let's see how the board looks with `r bin_width` bins:

```{r board_2, fig.height = 5, fig.width = 1.5}
board = plinko_board(ev_dist, bin_width = bin_width)

autoplot(board, show_paths = FALSE) +
  geom_vline(xintercept = 269)
```

That looks pretty good! If we want to get *really* persnickety about it, we
are still off by a little over 1 electoral vote:

```{r closest_edge_at_bin_width}
closest_edge_to_269(bin_width)
```

We could fix that by shifting the center of the board slightly to compensate. To
do that we need to know the signed difference between the slot edge closest to
269 and 269:

```{r slot_edge_shift}
slot_edge_shift = 269 - board$slot_edges[which.min(abs(board$slot_edges - 269))]
slot_edge_shift
```

Then we can use this to shift the board center slightly

```{r board_3, fig.height = 5, fig.width = 1.5}
shifted_center = board$center + slot_edge_shift
board = plinko_board(ev_dist, bin_width = bin_width, center = shifted_center)

autoplot(board, show_paths = FALSE) +
  geom_vline(xintercept = 269)
```


## Customizing board appearance

Now that our board has the parameters we want---a distribution that matches
our target distribution and a slot edge falling on 269---we can customize the
rest of the appearance of the board. 

### Adjusting limits

First, we can adjust the limits of the board using the `limits` parameter.
For this use case we should show the distribution compared to the full
range of possible electoral votes:

```{r board_limits, fig.height = 5, fig.width = 2}
board = plinko_board(ev_dist, bin_width = bin_width, center = shifted_center, limits = c(0, 538))

autoplot(board)
```

### Customizing ggplot layers

We can also customize
existing layers in the `ggplot` objects generated for 
each frame using the `modify_layer()` function, which takes a layer name
followed by parameters you would normally pass to a ggplot layer/geom.
Aesthetic mappings you provide are merged with the existing aesthetics in the
layer, and new arguments you provide override existing ones.

For example, say you want to make the ball fill color depend on its x position, and
change the outline color to be black:

```{r fig.height = 5, fig.width = 2.5}
board %>%
  modify_layer("balls", aes(fill = x), color = "black") %>%
  autoplot(show_dist = FALSE)
```

You can use `modify_layer()` to adjust the following layers:

- "slot_edges": A `geom_segment()` that draws the edges of the slots
- "pins": A `geom_point()` that draws the pins
- "paths": A `geom_path()` that draws ball paths
- "balls": A `geom_circle()` that draws the balls
- "dist": A `geom_step()` that draws the reference binomial distribution

When plotting a single frame, you can also add additional ggplot objects after
the call to `autoplot()`, as with any ggplot object:

```{r fig.height = 5, fig.width = 2}
board %>%
  autoplot(show_dist = FALSE, show_paths = FALSE) +
  geom_vline(xintercept = 269, color = "black", alpha = 0.15, size = 1) +
  annotate("label", x = 269, y = 1500, label = "269", hjust = 0.5, color = "gray50")
```

However, layers added in this way are not saved into the `board` object, and so 
**will not be displayed when the board is animated**.

To add ggplot objects to the `board` object so that they are displayed when the
board is animated, you must use the `add_layers()` function *before* calling
`autoplot()` or `animate()`. Here is the same example using `add_layers()`:

```{r fig.height = 5, fig.width = 2}
board %>%
  add_layers(
    geom_vline(xintercept = 269, color = "black", alpha = 0.15, size = 1),
    annotate("label", x = 269, y = 1500, label = "269", hjust = 0.5, color = "gray50")
  ) %>%
  autoplot(show_dist = FALSE, show_paths = FALSE, show_target_dist = FALSE)
```

As static plots, both look identical. However, when we get to animating (next), 
we'll have to use the `add_layers()` approach.

## Animating the board

By default, no tweening is done between balls on the board:

```{r animate_no_tween}
board %>%
  # width is determined automatically based on height
  animate(fps = 7.5, height = 500)
```

Most likely you will want some tweening between ball states. I find that using 4 times the
base number of frames with a `"bounce-out"` easing function (the default) makes for a smooth
animation that feels more physically accurate. You can add tweening using the
`tween_balls()` function. You'll also want to increase the `fps` to account for
the additional frames, and we will add a 3-second (90-frame) pause at the end 
of the animation using `end_pause`:

```{r animate_tween}
board %>%
  tween_balls(frame_mult = 4, ease = "bounce-out") %>%
  animate(fps = 30, height = 500, end_pause = 90)
```

For testing purposes (e.g. faster rendering), it can also be useful to filter the frames in the final animation.
You can do this using `filter_frames()`, which takes filtering conditions in the same format
as `filter()` and applies them to the data frame returned by `frames(board)`. The `"ball_id"` column
can be combined with the `"stopped"` column (which is `TRUE` after the ball hits the 
bottom of a slot and stops moving) to show just one ball dropping, for example:

```{r animate_one_ball}
board %>%
  filter_frames(ball_id == 1, !stopped) %>%
  tween_balls(frame_mult = 4, ease = "bounce-out") %>%
  animate(fps = 30, height = 500)
```

## A full, annotated board

Finally, we can combine everything together to adjust the existing geometries 
with `modify_layer()`, add annotations with `add_layers()`, add tweening
with `tween_balls()`, then animate:

```{r fig.height = 6, fig.width = 2.5}
Biden_color = "#0571b0"
Trump_color = "#ca0020"

annotation_height = 2000

board %>%
  modify_layer("pins", color = "gray50") %>%
  # the "balls" layer uses the frame(board) data frame, which has a "region"
  # column giving the part of the board the ball is in ("start", "pin", or "slot")
  # we can use this to change the ball color when it falls into the slot
  modify_layer(
    "balls", 
    aes(fill = ifelse(region == "slot", ifelse(x <= 269, "Trump", "Biden"), "none")),
    color = NA
  ) %>%
  add_layers(
    geom_vline(xintercept = 269, color = "black", alpha = 0.15, size = 1),
    annotate("text", 
      x = 290, y = 0.92 * annotation_height,
      label = "Biden wins", hjust = 0, color = Biden_color,
    ),
    annotate("text", 
      x = 250, y = 0.92 * annotation_height,
      label = "Trump wins", hjust = 1, color = Trump_color
    ),
    annotate("label", 
      x = 269, y = 0.97 * annotation_height,
      label = "269", hjust = 0.5, color = "gray50",
      fontface = "bold"
    ),
    expand_limits(y = annotation_height),
    scale_fill_manual(
      limits = c("none", "Biden", "Trump"),
      values = c("gray45", Biden_color, Trump_color), 
      guide = FALSE
    ),
    theme(axis.title.x = element_text(hjust = 0, size = 10, color = "gray25")),
    xlab("Electoral votes for Biden")
  ) %>%
  tween_balls(frame_mult = 4, ease = "bounce-out") %>%
  animate(fps = 30, height = 600, end_pause = 90)
```
