---
title: "Presidential Plinko"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{presidential_plinko}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style type="text/css">
.kable-table table {
  margin-left: 0;
}
img {
  border: none;
}
</style>

```{r chunk_options, include=FALSE}
if (capabilities("cairo") && Sys.info()[['sysname']] != "Darwin") {
  knitr::opts_chunk$set(
    dev.args = list(png = list(type = "cairo"))
  )
}
```


The following code constructs a Plinko board for a quantile dotplot of a
predictive distribution for the 2020 US Presidential election outcome.

See [presidential-plinko.com](https://presidential-plinko.com) and the 
[mjskay/election-galton-board](https://github.com/mjskay/election-galton-board)
repository for more information.

## Setup

The following packages are required for this example:

```{r setup, message=FALSE, warning=FALSE}
library(plinko)
library(dplyr)
library(ggplot2)
library(ggdist)
library(ggforce)

theme_set(theme_ggdist())
```

## Desired target sample

Say we've taken 50 quantiles from a predictive distribution for the number of 
electoral votes Biden will receive in the 2020 US Presidential election:

```{r}
# this 20 quantiles from the Economist's forecast as of Sept 29, 2020.
x = c(213, 243, 266, 279, 290, 304, 310, 319, 323, 333, 339, 345,  352, 357, 363, 372, 374, 385.125, 399, 415)

tibble(x) %>%
  ggplot(aes(x = x)) +
  geom_dots() +
  xlab("Predicted electoral votes for Biden") +
  xlim(0, 538)
```

What we would like to do is construct a Plinko board that "sediments" into this dotplot. That is,
it gives us an animation of a plausible random process that could have led to a dotplot
that looks like the one above. This will take several steps.

## Constructing the board

First, to determine the board dimensions, we need to decide on the number of draws we will 
take and the number of pins in one row of the board (equivalently, the size of the Binomial 
distribution we are simulating with the board):

```{r}
n_bin = 6
bin_width = 41
board_center = 330.5
```

Then we can construct the Plinko board:

```{r}
set.seed(12345)

board = plinko_board(
  x, n_bin, bin_width, center = board_center,
  limits = c(0, 538)#, total_height = 2445
)

board
```

## Plotting the board

### Using `autoplot()`

The Plinko board will automatically find random paths the balls could have taken
to end up in their final locations. We can see these paths by plotting the board
with the `autoplot()` function, which returns a `ggplot()` object:

```{r fig.height = 4.5, fig.width = 2.1}
board %>%
  autoplot()
```

By default the plot overlays the Binomial distribution implied by the Plinko board so 
you can judge the quality of the approximation. We can also turn this off:

```{r fig.height = 4.5, fig.width = 2.1}
board %>%
  autoplot(show_dist = FALSE)
```

You can also see the board without paths by passing `show_paths = FALSE`, and you 
can plot specific frames of the animation by passing the `frame` parameter:

```{r fig.height = 4.5, fig.width = 2.1}
autoplot(board, show_paths = FALSE, frame = 9)
```

### Plotting manually

We could also create a plot like the ones above manually by using the `slot_edges()`, `pins()`, `paths()`, and `balls()`
unctions, which return data frames containing the locations of all of the elements of the board:

```{r board_plot, fig.height = 4.5, fig.width = 2.1}
ggplot() + 
  geom_segment(aes(x = x, y = 0, xend = x, yend = height), data = slot_edges(board), color = "gray75", size = 1) + 
  geom_point(aes(x, y), data = pins(board), shape = 19, color = "#e41a1c", size = 1) +
  geom_path(aes(x = x, y = y, group = ball_id), data = paths(board), alpha = 1/4, size = 1, color = "gray50") +
  geom_circle(aes(x0 = x, y0 = y, r = width/2), data = balls(board), fill = "#1f78b4", color = NA) +
  coord_fixed(expand = FALSE, clip = "off") +
  ylab(NULL) +
  scale_y_continuous(breaks = NULL) +
  theme(
    axis.line.y = element_blank(),
    axis.line.x = element_line(color = "gray75", size = 1)
  ) 
```

However, if you wish to customize the plot, it is probably easier to use the
`modify_layer()` and `add_layers()` functions described in the next section, as
these also impact the frames used when an animated plot is rendered.

## Customizing frames

You can customize the existing layers in the `ggplot` objects generated for 
each frame using the `modify_layer()` function, which takes a layer name
followed by parameters you would normally pass to a ggplot layer/geom.
Aesthetic mappings you provide are merged with the existing aesthetics in the
layer, and new arguments you provide override existing ones.

For example, say you want to make the ball fill color depend on its x position, and
change the outline color to be black:

```{r fig.height = 4.5, fig.width = 3}
board %>%
  modify_layer("balls", aes(fill = x), color = "black") %>%
  autoplot(show_dist = FALSE)
```

You can use `modify_layer()` to adjust the following layers:

- "slot_edges": A `geom_segment()` that draws the edges of the slots
- "pins": A `geom_point()` that draws the pins
- "paths": A `geom_path()` that draws ball paths
- "balls": A `geom_circle()` that draws the balls
- "dist": A `geom_step()` that draws the reference binomial distribution

When plotting a single frame, you can also add additional ggplot objects after
the call to `autoplot()`, as with any ggplot object:

```{r fig.height = 5, fig.width = 2.1}
board %>%
  autoplot(show_dist = FALSE, show_paths = FALSE) +
  geom_vline(xintercept = 269, color = "black", alpha = 0.15, size = 1) +
  annotate("label", x = 269, y = 1500, label = "269", hjust = 0.5, color = "gray50")
```

However, layers added in this way are not saved into the `board` object, and so 
**will not be displayed when the board is animated**.

To add ggplot objects to the `board` object so that they are displayed when the
board is animated, you must use the `add_layers()` function *before* calling
`autoplot()` or `animate()`. Here is the same example using `add_layers()`:

```{r fig.height = 5, fig.width = 2.1}
board %>%
  add_layers(
    geom_vline(xintercept = 269, color = "black", alpha = 0.15, size = 1),
    annotate("label", x = 269, y = 1500, label = "269", hjust = 0.5, color = "gray50")
  ) %>%
  autoplot(show_dist = FALSE, show_paths = FALSE)
```

As static plots, both look identical. However, when we get to animating (next), 
we'll have to use the `add_layers()` approach.

## Animating the board

By default, no tweening is done between balls on the board:

```{r}
board %>%
  # width is determined automatically based on height
  animate(fps = 7.5, height = 500)
```

Most likely you will want some tweening between ball states. I find that using 4 times the
base number of frames with a `"bounce-out"` easing function (the default) makes for a smooth
animation that feels more physically accurate. You can add tweening using the
`tween_balls()` function. You'll also want to increase the `fps` to account for
the additional frames, and we will add a 3-second (90-frame) pause at the end 
of the animation using `end_pause`:

```{r}
board %>%
  tween_balls(frame_mult = 4, ease = "bounce-out") %>%
  animate(fps = 30, height = 500, end_pause = 90)
```

For testing purposes (e.g. faster rendering), it can also be useful to filter the frames in the final animation.
You can do this using `filter_frames()`, which takes filtering conditions in the same format
as `filter()` and applies them to the data frame returned by `frames(board)`. The `"ball_id"` column
can be combined with the `"stopped"` column (which is `TRUE` after the ball hits the 
bottom of a slot and stops moving) to show just one ball dropping, for example:

```{r}
board %>%
  filter_frames(ball_id == 1, !stopped) %>%
  tween_balls(frame_mult = 4, ease = "bounce-out") %>%
  animate(fps = 30, height = 500)
```

## A full, annotated board

Finally, we can combine everything together to adjust the existing geometries 
with `modify_layer()`, add annotations with `add_layers()`, add tweening
with `tween_balls()`, then animate:

```{r fig.height = 5, fig.width = 5}
Biden_color = "#0571b0"
Trump_color = "#ca0020"

annotation_height = 1400

board %>%
  modify_layer("pins", color = "gray50") %>%
  # the "balls" layer uses the frame(board) data frame, which has a "region"
  # column giving the part of the board the ball is in ("start", "pin", or "slot")
  # we can use this to change the ball color when it falls into the slot
  modify_layer(
    "balls", 
    aes(fill = ifelse(region == "slot", ifelse(x <= 269, "Trump", "Biden"), "none")),
    color = NA
  ) %>%
  add_layers(
    geom_vline(xintercept = 269, color = "black", alpha = 0.15, size = 1),
    annotate("text", 
      x = 290, y = 0.92 * annotation_height,
      label = "Biden wins", hjust = 0, color = Biden_color,
    ),
    annotate("text", 
      x = 250, y = 0.92 * annotation_height,
      label = "Trump wins", hjust = 1, color = Trump_color
    ),
    annotate("label", 
      x = 269, y = 0.97 * annotation_height,
      label = "269", hjust = 0.5, color = "gray50",
      fontface = "bold"
    ),
    expand_limits(y = annotation_height),
    scale_fill_manual(
      limits = c("none", "Biden", "Trump"),
      values = c("gray45", Biden_color, Trump_color), 
      guide = FALSE
    ),
    theme(axis.title.x = element_text(hjust = 0, size = 10, color = "gray25")),
    xlab("Electoral votes for Biden")
  ) %>%
  tween_balls(frame_mult = 4, ease = "bounce-out") %>%
  animate(fps = 30, height = 500, end_pause = 90)
```
